|

### 2021-05-30 article/article.Rmd
add discussion

---
title:  |
        | The Effect of Legislature Size on Public Spending:
        | A Meta-Analysis
author: 
  - Huzeyfe Alptekin[^huzeyfe]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
  - Catarina Roman[^catarina]
date: \today
thanks: "The authors thank Guilherme Duarte, Robert Myles McDonnell, and David Skarbek for their constructive feedback. We also thank Cedric Antunes, Luis Castro, Giovanna França, Julia Oriente, and Lucas Mingardi for their excellent research assistance. Replication materials are available at <https://github.com/danilofreire/legislature-size-meta-analysis>. We kindly acknowledge funding from the São Paulo State Science Foundation (FAPESP grant number 2018/00646-1)."
abstract: "In a seminal article, @weingast1981political argue that there is a positive relationship between legislature size and inefficiency in public expenditures. Their proposition is currently known as the \"law of $1/n$\" and has been widely debated in political science and public administration. However, recent studies have questioned the validity of the theory. In this letter, we estimate the first meta-analysis of the relationship between the number of legislators and public spending. Based on a sample of 29 articles, we find no robust evidence for the effect of legislature size on government budgets. Yet the aggregate results show significant heterogeneity. While earlier studies provide moderate support for the \"law of $1/n$\", papers using causal inference methods consistently find a negative relationship between seats and spending. The available evidence also indicates that proportional representation and mixed voting systems are no more likely to overspend than majoritarian ones."
abstractspacing: double
keywords: distributive politics; law of $1/n$; legislature size; meta-analysis; public spending
jelcodes: H21; H23; H50; H61
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

In the second round, we did not require articles to cite @weingast1981political
and used a keyword-based query on Google Scholar to broaden the scope of the
first search. The search string contained terms strongly associated with the
literature on the "law of $1/n$" and was as follows: `("upper chamber size" OR
"lower chamber size" OR "council size" OR "parliament size" OR "legislature
size" OR "number of legislators" OR "legislative size") AND ("spending" OR
"expenditure" OR "government size")`. We again restricted the search to
articles written in English which employed quantitative methods. This search
added two new results to our sample [@coate2011government;
@debenedetto2018effect], but neither of them included variables beyond the six
measures we had previously identified. In the third search round, we looked
into the personal webpages of every author whom we had already included in our
sample. The purpose of this manual search was to assess whether there was any
working paper or unpublished manuscript that we had missed in the two former
queries. All papers we found in the last search had already been included in
our sample. Combined, the three searches produced a dataset of 
`r length(unique(dat$id))` studies as of the 10\textsuperscript{th} of March 2021.
Table \ref{tab:papers} contains the full list of articles we analyse in this
paper.

@@

The uppermost row shows the results for lower chamber size in our restricted
sample. In the first model, which correlates lower house and expenditure per
capita, we find a standardised mean difference (SMD) of 0.022 and a standard
error of 0.131 (studies = `r mod[[1]]$k`, 95\% CI = [-0.256; 0.299], $p$-value
= 0.87), so we cannot rule that the effect is different from zero. Indeed, the
effect of lower chamber size on the other two dependent variables is also null
in statistical terms. When we compare lower chamber size with log expenditure
per capita, the overall effect size is -0.031 and the standard error is 0.049
(studies = `r mod[[3]]$k`, 95\% CI = [-0.188; 0.127], $p$-value = 0.58), and
the impact of larger lower houses on government spending as a percentage of GDP
is also negligible (studies = `r mod[[5]]$k`, SMD = - 0.006, 95\% CI =
[-0.0334; 0.021], $p$-value = 0.563). The results are virtually identical when
we estimate the meta-analyses using our extended sample, and the all three
coefficients are again statistically indistinguishable from zero.

Next, we present the meta-analyses using the logarithm of lower house size as
an independent variable. The relationship between this variable and the log of
per capita expenditure is positive, but the coefficient is not significant
(studies = `r mod[[4]]$k`, SMD = 0.078, SE = 0.109, 95\% CI = [-0.225; 0.381],
$p$-value = 0.515. The effect of log lower house size on expenditure as a
percentage of the GDP is very similar, although the SMD is negative (studies =
`r mod[[6]]$k`, SMD = `r round(mod[[6]]$TE.random,3)`, SE = 
`r round(mod[[6]]$seTE.random,3)`, 95\% CI = [`r round(mod[[6]]$lower.random,3)`;
`r round(mod[[6]]$upper.random,3)`], $p$-value = 
`r round(mod[[6]]$pval.random,3)`). Results in the full sample are also null, and
the coefficients for each dependent variable have the same sign as the
restricted sample -- positive and negative, respectively.

The third set of models uses upper house size as the main independent variable.
We find a positive correlation between this variable and expenditure per capita
(studies = `r mod[[2]]$k`, SMD = 3.658, SE = 4.299, 95\% CI = [-6.255; 13.572],
$p$-value = 0.419), and a negative relationship with government spending as a
percentage of GDP (studies = `r mod[[7]]$k`, SMD = 
`r round(mod[[7]]$TE.random,3)`, SE = `r round(mod[[7]]$seTE.random,3)`, 95\% CI =
[`r round(mod[[7]]$lower.random,3)`; `r round(mod[[7]]$upper.random,3)`],
$p$-value = `r round(mod[[7]]$pval.random,3)`), yet neither coefficient is
statistically significant. Results are the same in our extended sample.

@@

In this letter, we assess the empirical validity of the "law of $1/n$". Based
on a sample of `r length(unique(dat$id))` publications on the topic, our
meta-analyses show that there is no strong evidence that an increase in the
number of legislators has a significant effect on public expenditures. If such
effect exists, it is likely driven by an increase in $k$, the size of the upper
legislature, as suggested by several studies in the literature
[@baqir2002districting; @bradbury2001legislative; @bradbury2003local;
@chen2007law; @gilligan2001fiscal; @primo2006stop]. Instead, we find
better evidence for the "reverse law of $1/n$", which posits that larger
legislatures lead to lower government spending. This is mainly because studies
using regression discontinuity designs, a method that has robust internal
validity, consistently indicate a negative relationship between lower house size
and the logarithm of expenditure per capita [@hohmann2017effect;
@lewis2019legislature; @petterssonlidbom2012size].

The meta-regressions show that study characteristics have a considerable
influence on reported results. Electoral system affects the relationship
between legislature size and public expenditure, but the results are not
replicable in all estimations. Publication year generates conflicting findings
in our models. Nevertheless, the meta-regressions confirm that RDDs produce
negative effects more frequently than OLS regressions.

Why is there no clear-cut evidence in favour or against the "law of $1/n$"? A
plausible reason may be that there are few incentives for the pure accumulation
of knowledge in the social sciences, at least when compared to the benefits
scholars may accrue when they challenge or add features to existing theories
[@geddes2003paradigms]. This leads to a reduced number of replication studies in
the field, although we have seen some positive changes in this respect, such as
EGAP's _Metaketa Initiative_.\footnote{See
\url{https://egap.org/our-work/the-metaketa-initiative} for further
information.} For instance, in our sample, papers added supermajority rules
[@lee2015supermajority; @lee2016supermajority], redistricting
[@baqir2002districting; @lee2018court], party ideology [@bjedov2014impact],
coalition sizes [@baskaran2013coalition], term limits [@erler2007termlimits],
bicameralism [@ricciuti2004legislature], and the interplay between upper and
lower houses [@chen2007law] to the main theory, but a comprehensive
procedural replication of the findings using different samples is yet to be
written. The addition of new features has the benefit of enriching the original
theory with useful details, although it has the disadvantage of not providing a
conclusive test to the "law of $1/n$".

Our analyses suggest three areas for further research. First, our study
sample did not include articles that evaluate the association between the
log($n$) and public expenditure per capita or between $k$ and log expenditure
per capita. New work on that area might clarify some of the inconsistencies we
find here. Second, despite the inclusion of several moderators in our models,
aggregate results still show considerable heterogeneity. Domestic factors such
as party dynamics or gerrymandering [@lee2015supermajority;
@mukherjee2003politicalparties; @gilligan2006public] may prove useful in this
regard. Finally, we highlight the need for more causal inference studies in the
literature. Whenever possible, authors should leverage natural and
quasi-experiments to assess whether the current results hold when tested with
such research designs. These suggestions may help scholars and policy-makers to
reach an optimal balance between sound fiscal policy and the demands for
increased political representation.

\newpage
\setlength{\parindent}{0cm}
\setlength{\parskip}{5pt}

\nocite{baskaran2013coalition, bradbury2009spatially, drew2017price,
erler2007termlimits, fiorino2007legislature, hohmann2017effect,
kessler2014communication, lewis2019legislature, lledo2003electoral,
mukherjee2003politicalparties, petterssonlidbom2012size, schaltegger2009large,
stein1998institutional, mukherjee2003politicalparties, macdonald2008impact,
matsusaka2005endogeneity, ricciuti2004legislatures}

>>>>

---
title:  |
        | The Effect of Legislature Size on Public Spending:
        | A Meta-Analysis
author: 
  - Huzeyfe Alptekin[^huzeyfe]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
  - Catarina Roman[^catarina]
date: \today
thanks: "The authors thank Guilherme Duarte, Robert Myles McDonnell, and David Skarbek for their constructive feedback. We also thank Cedric Antunes, Luis Castro, Giovanna França, Julia Oriente, and Lucas Mingardi for their excellent research assistance. Replication materials are available at <https://github.com/danilofreire/legislature-size-meta-analysis>. We kindly acknowledge funding from the São Paulo State Science Foundation (FAPESP grant number 2018/00646-1)."
abstract: "In a seminal article, @weingast1981political argue that there is a positive relationship between legislature size and inefficiency in public expenditures. Their proposition is currently known as the \"law of $1/n$\" and has been widely debated in political science and public administration. However, recent studies have questioned the validity of the theory. In this letter, we estimate the first meta-analysis of the relationship between the number of legislators and public spending. Based on a sample of 30 articles, we find no robust evidence for the effect of legislature size on government budgets. Yet the aggregate results show significant heterogeneity. While earlier studies provide moderate support for the \"law of $1/n$\", papers using causal inference methods consistently find a negative relationship between seats and spending. The available evidence also indicates that proportional representation and mixed voting systems are no more likely to overspend than majoritarian ones."
abstractspacing: double
keywords: distributive politics; law of $1/n$; legislature size; meta-analysis; public spending
jelcodes: H21; H23; H50; H61
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

In the second round, we did not require articles to cite @weingast1981political
and used a keyword-based query on Google Scholar to broaden the scope of the
first search. The search string contained terms strongly associated with the
literature on the "law of $1/n$" and was as follows: `("upper chamber size" OR
"lower chamber size" OR "council size" OR "parliament size" OR "legislature
size" OR "number of legislators" OR "legislative size") AND ("spending" OR
"expenditure" OR "government size")`. We again restricted the search to
articles written in English which employed quantitative methods. This search
added two new results to our sample [@coate2011government;
@debenedetto2018effect], but neither of them included variables beyond the six
measures we had previously identified. In the third search round, we looked
into the personal webpages of every author whom we had already included in our
sample. The purpose of this manual search was to assess whether there was any
working paper or unpublished manuscript that we had missed in the two former
queries. All papers we found in the last search had already been included in
our sample. The full list of excluded records is available for online
consultation in the replication materials. Combined, the three searches
produced a dataset of `r length(unique(dat$id))` studies as of the
10\textsuperscript{th} of March 2021. Table \ref{tab:papers} contains the full
list of articles we analyse in this paper.

@@

The uppermost row shows the results for lower chamber size in our restricted
sample. In the first model, which correlates lower house and expenditure per
capita, we find a standardised mean difference (SMD) of 0.022 and a standard
error of 0.131 (figure 1.1, studies = `r mod[[1]]$k`, 95\% CI = [-0.256;
0.299], $p$-value = 0.87), so we cannot rule that the effect is different from
zero. Indeed, the effect of lower chamber size on the other two dependent
variables is also null in statistical terms. When we compare lower chamber size
with log expenditure per capita, the overall effect size is -0.031 and the
standard error is 0.049 (figure 1.2, studies = `r mod[[3]]$k`, 95\% CI =
[-0.188; 0.127], $p$-value = 0.58), and the impact of larger lower houses on
government spending as a percentage of GDP  is also negligible (figure 1.3,
studies = `r mod[[5]]$k`, SMD = - 0.006, 95\% CI = [-0.0334; 0.021], $p$-value
= 0.563). The results are virtually identical when we estimate the
meta-analyses using our extended sample, and the all three coefficients are
again statistically indistinguishable from zero.

Next, we present the meta-analyses using the logarithm of lower house size as
an independent variable. The relationship between this variable and the log of
per capita expenditure is positive, but the coefficient is not significant
(figure 1.4, studies = `r mod[[4]]$k`, SMD = 0.078, SE = 0.109, 95\% CI =
[-0.225; 0.381], $p$-value = 0.515). The effect of log lower house size on
expenditure as a percentage of the GDP is negative, but it is again
non-statistically significant (figure 1.5, studies = `r mod[[6]]$k`, SMD = 
`r round(mod[[6]]$TE.random,3)`, SE = `r round(mod[[6]]$seTE.random,3)`, 95\% CI =
[`r round(mod[[6]]$lower.random,3)`; `r round(mod[[6]]$upper.random,3)`],
$p$-value = `r round(mod[[6]]$pval.random,3)`). Results in the full sample are
also null, and the coefficients for each dependent variable have the same sign
as the restricted sample -- positive and negative, respectively.

The third set of models uses upper house size as the main independent variable.
We find a positive correlation between this variable and expenditure per capita
(figure 1.6, studies = `r mod[[2]]$k`, SMD = 3.658, SE = 4.299, 95\% CI =
[-6.255; 13.572], $p$-value = 0.419), and a negative relationship with
government spending as a percentage of GDP (figure 1.7, studies = 
`r mod[[7]]$k`, SMD = `r round(mod[[7]]$TE.random,3)`, SE = 
`r round(mod[[7]]$seTE.random,3)`, 95\% CI = [`r round(mod[[7]]$lower.random,3)`;
`r round(mod[[7]]$upper.random,3)`], $p$-value = 
`r round(mod[[7]]$pval.random,3)`), yet neither coefficient is statistically
significant. Results are the same in our extended sample.

@@

In this letter, we use meta-analytic methods to assess the generality of the
"law of $1/n$". Based on a sample of `r length(unique(dat$id))` publications,
our meta-analyses show that there is no strong evidence that an increase in the
number of legislators has a significant effect on public expenditures. The
meta-regressions indicate that study characteristics have a considerable
influence on reported results. Electoral system affects the relationship
between legislature size and public expenditure, but the results are not
replicable in all estimations. In our extended sample, we find that
unicameralism favours the $1/n$ effect. Publication year generates conflicting
findings in our models, but recent papers tend to present smaller coefficients
than older ones. The meta-regressions confirm that modern estimation methods
such as RDDs and panel/fixed-effects models decrease effects more frequently
than OLS regressions. In a nutshell, we only find moderate support for the "law
of $1/n$" in unicameral governments and in the upper house, mostly when authors
used OLS estimators.

While the vast literature covering the "law of $1/n$" builds important
empirical knowledge, we hypothesise that some of the null findings that we
present here are due to difficulties in testing important assumptions behind
the theory itself. For instance, the theory assumes three types of costs for
legislative public good provision, namely expenses for the constituency,
expenses outside the constituency, and externalities. The main issue in
assessing their actual impact is that externalities such as shifts in prices of
local firms, for example, are extremely hard to measure. Thus, it is very
difficult to properly translate the mechanism to empirical data, making it is
easy to accidentally distort results. Therefore, it should not come as a
surprise that slight differences in political features generate highly
heterogeneous results.

In this sense, the empirical cases in the literature may not always be the most
fortunate testing ground for the "law of $1/n$". While we believe that moving
beyond the majoritarian districts framework could produce valuable insights,
institutional features that are central to the theory cannot be overrun. For
example, proportional representation (PR) electoral systems allow candidates
whose constituents are spread across large territories to provide diffused
public goods and win elections. However, geographically-targeted service
provision is at the very core of the legislative behaviour that produces the
"law of $1/n$". Thus, scholars should consider the possible implications of
these micro-level dynamics when applying the "law of $1/n$" logic to different
settings.

Another plausible reason why there is no clear-cut evidence in favour or
against the "law of $1/n$" may be that there are few incentives for the pure
accumulation of knowledge in the social sciences, at least when compared to the
benefits scholars may accrue when they challenge or add features to existing
theories [@geddes2003paradigms]. This leads to a reduced number of replication
studies and research syntheses in the field, although we have seen some
positive changes in this respect, such as EGAP's _Metaketa
Initiative_.\footnote{See
\url{https://egap.org/our-work/the-metaketa-initiative} for further
information.} Here we show that continuously producing and consuming
meta-analyses is a viable path towards knowledge-building in our discipline.
Although aggregating observational -- instead of experimental -- data can be
especially challenging, we believe that research synthesis methods play an
essential role in advancing our understanding of complex political phenomena.

Our analyses suggest other avenues for further research. First, our study
sample did not include articles that evaluate the association between the
natural logarithm of lower chamber size and public expenditure per capita, or
between upper house size and log expenditure per capita. New work on that area
might clarify some of the inconsistencies we find here. Second, despite the
inclusion of several moderators in our models, aggregate results still show
considerable heterogeneity. Domestic factors such as party dynamics or
gerrymandering [@lee2015supermajority; @mukherjee2003politicalparties;
@gilligan2006public] may prove useful in explaining those divergent results.
Third, authors should leverage natural and quasi-experiments to assess whether
the current results hold when systematically tested. Fortunately, this may also
be a trend under way, as all four studies using regression discontinuity
designs in our sample were published within less than 10 years prior to this
meta-analysis. These suggestions may help scholars to validate the robustness
of their findings and policy-makers to reach an optimal balance between sound
fiscal policy and the demands for increased political representation.

\setlength{\parindent}{0cm}
\setlength{\parskip}{5pt}

\nocite{baskaran2013coalition, bradbury2009spatially, drew2017price,
erler2007termlimits, fiorino2007legislature, hohmann2017effect,
kessler2014communication, lewis2019legislature, lledo2003electoral,
mukherjee2003politicalparties, petterssonlidbom2012size, schaltegger2009large,
stein1998institutional, mukherjee2003politicalparties, macdonald2008impact,
matsusaka2005endogeneity, ricciuti2004legislatures, coate2011government}
